Easy as Pi: A Network Coding Raspberry Pi Testbed

### 2016-06-06 testbed.tex
began persistant storage
first draft of testbed ready

\subsection{Write image}

The final step is to write the image to all the memory cards before they can
be plugged in the \ac{Raspi}s. This is done by first leaving the chrooted filesystem
and unmounting all the directories we mounted:

% Exit chroot, umount, and write to memory card
\begin{lstlisting}[]
# exit
# cd ..
# umount --recursive ${ROOTDIR}
# losetup -d WHICH_NUMBE.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

If successfully unmounted, you can insert a memory card in your machine. In our
system, this memory card was available under /dev/mmcblk0, but this may be
different on different machines. For internal memory card readers, it will
probably be named /dev/mmcblkX where X is some number and if the memory card
reader is connected through \ac{USB}, then it is likely called /dev/sdX.
Make sure to write to the correct device as everything will be overwritten!

Below is a few commands that may help you to deduce which device is correct:

@@

For us, the device was /dev/mmcblk0, so we write the image to that device:
\begin{lstlisting}[]
# dd if=${IMAGE}.img of=/dev/mmcblk0 bs=4M
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

>>>>

At this point, it is no longer required to be \texttt{chrooted} and the image
can be written to a memory card. The following commands exits the \texttt{chrooted}
environment, unmount all partitions, and detaches the loop device:

% Exit chroot, umount, and write to memory card
\begin{lstlisting}[]
# exit
# cd ..
# umount --recursive ${ROOTDIR}
# losetup -d ${DEV}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\subsection{(Optional) Persistent subdirectories}

Provided the stacked filesystem was configured, it may be desired to have
some directories where files are not flushed upon rebooting the \ac{Raspi}s.
The following procedure will create an extra partition in the image for
the \ac{Raspi} users home directory that will be made storage persistent.
%Provided that a stacked filesystem was configured to make the root
%filesystem forget chances upon reboots is not always desired.

Expand image according to the desired home directory size. Keep in mind not
to make the image bigger than the target memory card.
% Expand image
\begin{lstlisting}[]
# dd if=/dev/zero bs=1M count=1024 >> ${IMAGE}.img && sync
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Create home partition after the root partition. Use \texttt{fdisk} to
find the next available sector in the image. In our case, the new
partition should start at sector 4806656:
\begin{lstlisting}[]
# fdisk -lu ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 3.3 GiB, 3534749696 bytes, 6903808 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

Device                               Boot  Start     End Sectors  Size Id Type
2016-05-27-raspbian-jessie-lite.img1        8192  137215  129024   63M  c W95 FAT32 (LBA)
2016-05-27-raspbian-jessie-lite.img2      137216 4806655 4669440  2.2G 83 Linux
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Create new partition:

\begin{lstlisting}[]
# (echo n; echo p; echo ; echo 4806656; echo ; echo p; echo w) | fdisk ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Create loopback device: 
\begin{lstlisting}[]
# DEV=$(losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Format new partition:
\begin{lstlisting}[]
# mkfs.ext4 ${DEV}p3
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The last thing to remember is to inform Raspbian to mount the home partition
we created during startup. This can be done by adding an entry in \texttt{fstab}:
\begin{lstlisting}[]
# mount ${DEV}p2 ${ROOTDIR}
# echo "/dev/mmcblk0p3 /home ext4 defaults,noatime 0 2" >> ${ROOTDIR}/etc/fstab
# umount ${ROOTDIR}
# losetup -d ${DEV}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\subsection{Write image to memory card}

The final step is to write the image to all the memory cards before they can
be used in the \ac{Raspi}s.

%This is done by first leaving the \texttt{chrooted}
%filesystem and unmounting all the directories we mounted:

% Exit chroot, umount, and write to memory card
%\begin{lstlisting}[]
%# exit
%# cd ..
%# umount --recursive ${ROOTDIR}
%# losetup -d ${DEV}.img
%\end{lstlisting}
%\FloatBarrier
%\vspace{-5mm}

%If successfully unmounted, you can 
Insert a memory card in your machine. It will be available as
\texttt{/dev/mmcblkX} or \texttt{/dev/sdX}

%In most systems, this memory card was available under \texttt{/dev/mmcblk0}, but this may be
%different on different machines. For internal memory card readers, it will
%probably be named /dev/mmcblkX where X is some number and if the memory card
%reader is connected through \ac{USB}, then it is likely called /dev/sdX.
Make absolutely sure to write to the correct device as everything will be 
overwritten!
Inserting/Ejecting the memory card a couple of times and using the commands
below may help deducing the correct device:
%The commands below may help deduce which device is the correct:
%Below is a few commands that may help you to deduce which device is correct:

@@

For us, the device was \texttt{/dev/mmcblk0}, so we write the image to that device:
\begin{lstlisting}[]
# dd if=${IMAGE}.img of=/dev/mmcblk0 bs=4M
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
